---
title: 'MECH481A6: Engineering Data Analysis in R'
subtitle: 'Chapter 5 Homework: Exploring Univariate Data' 
author: 'Student Name'
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: pdf_document
---

```{r global-options, include = FALSE}
# set global options for figures, code, warnings, and messages
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = "../figs/",
                      echo = FALSE, warning = FALSE, message = FALSE)
```

# Grading

We will grade the **knitted** PDF or HTML document from within your private
GitHub repository. Remember to make regular, small commits (e.g., at least one
commit per question) to save your work. We will grade the latest knit, as long
as it occurs *before* the start of the class in which we advance to the next
chapter. As always, reach out with questions via GitHub Issues or during
office hours.

# Data

You are probably sick of seeing the ozone data, but there's still more to
do with the file. Ozone concentration measurement is considered univariate,
thus we can use basic exploratory data analysis approaches to examine the data.

# Preparation

Load the necessary R packages into your R session. 

```{r packages, message=FALSE}
# load packages

library("dplyr")
library("ggplot2")
library("readr")
library("knitr")
library("kableExtra")
```

Recreate the pipe of `dplyr` functions that you used to import the data, select
and rename the variables listed below, drop missing observations, and assign the
output with a good name.

- `sample_measurement` renamed as `ozone_ppm` (ozone measurement in ppm)
- `datetime` (date in YYYY-MM-DD format and time of measurement in HH:MM:SS)

```{r import-data, message=FALSE}
# ozone: import, select, drop missing observations, rename

ppm_and_t <- read_csv("/Users/thomasrobert08/Documents/University /Colorado State University /Courses/Senior/Fall '25/Data Analysis in R/homework/ozone.csv") %>% 
  dplyr::select(sample_measurement, datetime) %>% 
  filter(!is.na(sample_measurement), 
         !is.na(datetime)) %>% 
  rename(ozone_ppm = sample_measurement)

```

Check that the data imported correctly.

```{r data-check}
# examine dataframe object 

ppm_and_t

```

\newpage

# Chapter 5 Homework: Exploring Univariate Data

Through Question 5, you will use all of the available ozone measurements
from January 2019 through January 2020. Starting in Question 6, you will use
a subset of the dataset: ozone concentration measurements on July 4, 2019.

## Question 1: Definitions

Guess the location, dispersion, and shape of ozone concentration data, based on
the definitions of each described in the coursebook. No code needed; just use
your intuition. For shape, take a look at the coursebook appendix on reference
distributions.

## Question 2: Quartiles

Calculate the quartiles of `ozone_ppm`. What is the minimum? Maximum? Median? 

```{r ozone-quartiles}
# calculate quantiles of ozone concentration

summary(ppm_and_t)

```

### Extra Credit

Create [a similar table](https://smogdr.github.io/edar_coursebook/eda1.html#descriptive-statistics) for `ozone_ppm`. Hint: You will need to investigate table
options in the `knitr` package.

\newpage

## Question 3: Cumulative Distribution Plot

Using either relevant `ggplot2` `geom` option, create a cumulative distribution
plot of `ozone_ppm`. Tweak the axis ranges for optimal data representation,
using `scale_*_continuous()` with `breaks =` and `minor_breaks =` arguments.
Add axis labels, title, subtitle, and theme. 

```{r ozone-cd}
# plot cumulative distribution of ozone concentration

ppm_and_t %>% 
  ggplot(aes(x = ozone_ppm)) + 
  stat_ecdf(geom = "step") + 
  scale_x_continuous(
    breaks = seq(0.00, 0.04, by = 0.005),
    minor_breaks = seq(0.00, 0.04, by = 0.001)
  ) +
  scale_y_continuous(
    breaks = seq(0, 1, by = 0.02),
    minor_breaks = seq(0, 1, by = 0.05)
  ) +
  labs(
    title = "Ozone Data Cumulative Distribution",
    subtitle = "Jan 2019 - Jan 2020", 
    x = "Ozone measurement (ppm)", 
    y = "Cumulative Proportion"
    ) + 
  theme_light() 

```

\newpage

## Question 4: Histogram

Create a histogram of `ozone_ppm`. Within the `geom`, mess with the number of
bins (e.g., 20, 50, 75, 100, 200) to explore the true shape and granularity of
the data. Match the plot style (e.g., title, subtitle, axis labels, theme) you
chose in Question 3, with the relevant adjustments such as "Histogram" 
instead of "Cumulative Distribution Plot".

```{r ozone-hist}
# create histogram of ozone concentration

ppm_and_t %>%
  ggplot(aes(x = ozone_ppm)) + 
  geom_histogram(bins = 50, fill = "lightblue", color = "black") +
  labs(
    title = "Ozone Data Histogram",
    subtitle = "Jan 2019 - Jan 2020",
    x = "Ozone Measurement (ppm)",
    y = "Count"
  ) +
  theme_light()

```

### Question 5: Concept

What mathematical concept is a histogram (Q4) attempting to visualize? 

Visualizes a distribution of data

### Question 6: Distribution

Based on the histogram (Q4), does ozone concentration appear to be normally
distributed? 

I would say that for the most part, it is. A good amount of the data is at a zero ppm value, but if one disregards this, the rest of the histogram is quite normally distributed. And this makes sense logically as well. 

### Question 7: Outliers

Based on the histogram (Q4), do you see any possible outliers? Skewness? How
might this affect the spread and central tendency?

As I mentioned in question 6, we see a few values at the zero ppm measurement. This skews the data, but makes sense logically. The central tendency in this case shifts towards lower amounts. 

\newpage

## Question 8: Boxplot

Generate a boxplot of ozone concentration on the y-axis with a title, subtitle,
y-axis label, and theme consistent with the style of the previous two plots. 
Use quotes (`""`) as the `x` arguments within the calls to the aesthetic and 
labels to remove the x-axis scale and label.

```{r ozone-boxplot}
# create ozone boxplot

ppm_and_t %>%
  ggplot(aes(x= "", y = ozone_ppm)) +
  geom_boxplot(fill = "salmon", color = "black") +
  labs(
    title = "Ozone Data Boxplot", 
    subtitle = "Jan 2019 - Jan 2020", 
    y = "Ozone Measurement (ppm", 
    x = ""
  ) +
  theme_light()
  )

```

\newpage

# Subset Data

Use the following code to create a dataframe for use in the remaining questions.
These ozone concentration measurements were taken on July 4, 2019 in Fort
Collins, CO. This code detects certain characters with the `datetime` object and
filters to observations containing those characters. There are other ways this 
could have been done (e.g., `dplyr::filter()` with `%in%` operator).

```{r ozone-day}
# create subset of data with only one day to examine daily pattern
# I did not ask you to code this because we have not discussed dates or stringr
# You need to uncomment the below three lines and run it; check object names
 ozone_day <- ppm_and_t %>% 
   dplyr::filter(stringr::str_detect(string = datetime,
                                     pattern = "2019-07-04"))
```

## Question 9: Autocorrelation Plot

Define autocorrelation as it relates to ozone concentration measurement. 

autocorrelation measures how much the current value of a variable is linearly related to its past values

Create an autocorrelation plot of ozone concentration, using `stats::acf()` and
include axis labels and title. Describe what you see based on the features of
interest outlined in the coursebook.

```{r ozone-acf}
# create autocorrelation plot with ozone_day df

stats::acf(
  x = ozone_day$ozone_ppm,
  main = "Autocorrelation of Ozone Concentration (July 4, 2019)", 
  xlab = "Lag (Hours)", 
  ylab = "Autocorrelation Function (ACF)"
)

```

\newpage

## Question 10: Parial Autocorrelation Plot

Define partial autocorrelation as it relates to ozone concentration measurement.

partial autocorrelation as it relates to ozone concentration measurements tells us the relationship between two measurements made close to each other in time

Now create a partial autocorrelation plot of day ozone concentration with axis
labels. Describe what you see. How does this compare to the autocorrelation
plot in the previous question?

```{r ozone-part-acf}
# create partial autocorrelation plot

stats::pacf(
  x = ozone_day$ozone_ppm,
  main = "Partial Autocorrelation of Ozone Concentrations (July 4, 2019)",
  xlab = "Lag (Hours)",
  ylab = "Partial Autocorrelation FUnction (PACF)"      
)

```

\newpage

# Appendix

```{r show_code, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}
```
